Weighted Model Counting with Conditional Weights for Bayesian Networks

### 2020-11-28 paper2/paper.tex
Fix two bugs in d-DNNF inference
Add TODOs to the paper
Remove half of the references

\section{Introduction}

Weighted model counting (WMC), i.e., an extension of model counting ($\#\SAT{}$)
that assigns a weight to every model \cite{DBLP:conf/aaai/SangBK05}, has emerged
as one of the most dominant and competitive approaches for handling inference
tasks in a wide range of formalisms including Bayesian networks
\cite{DBLP:conf/aaai/SangBK05,DBLP:books/daglib/0024906}, probabilistic
graphical models more generally \cite{DBLP:conf/ecsqaru/ChoiKD13}, and
probabilistic programs
\cite{DBLP:journals/tplp/FierensBRSGTJR15,DBLP:journals/corr/abs-2005-09089}.
Over the last fifteen years, WMC has been extended and generalised in many ways,
e.g., to handle continuous probability distributions
\cite{DBLP:conf/ijcai/BellePB15}, first-order probabilistic theories
\cite{DBLP:conf/ijcai/BroeckTMDR11,DBLP:journals/cacm/GogateD16}, and infinite
domains \cite{DBLP:conf/aaai/Belle17}. Furthermore, by generalising the notion
of weights to an arbitrary semiring, a range of other problems are also captured
\cite{DBLP:journals/japll/KimmigBR17}. Exact WMC solvers typically rely on
either knowledge compilation
\cite{DBLP:conf/ijcai/OztokD15,DBLP:conf/ijcai/LagniezM17} or exhaustive DPLL
search \cite{DBLP:conf/aaai/SangBK05}, whereas approximate solvers work by
sampling \cite{DBLP:conf/aaai/ChakrabortyFMSV14} and performing local search
\cite{DBLP:conf/sat/WeiS05}.

@@

\paragraph{ADDs and their use in inference.} ADDs provide an efficient way to
manipulate functions from a Boolean algebra to any algebraic structure (most
commonly real numbers) \cite{DBLP:journals/fmsd/BaharFGHMPS97}. They have been
used to represent value functions in Markov decision processes (MDPs)
\cite{DBLP:conf/uai/HoeySHB99} and extended to handle first-order MDPs
\cite{DBLP:journals/ai/SannerB09} and discrete and continuous state MDPs
\cite{DBLP:conf/uai/SannerDB11}. The latter extension has also been used for
weighted model integration---an extension of WMC that can handle continuous
variables \cite{DBLP:conf/ijcai/KolbMSBK18}. For Bayesian networks, ADDS have
been used to represent each conditional probability table (CPT) as an ADD
\cite{DBLP:conf/icml/ZhaoMP15} and as a combination of ADDs and arithmetic
circuits \cite{DBLP:conf/ijcai/ChaviraD07}. ADDs are particularly advantageous
in this situation because of their ability to fully exploit context-specific
independence, i.e., the observable structure within a CPT that is not inherited
from the structure of the Bayesian network \cite{DBLP:conf/uai/BoutilierFGK96}.

@@

We also consider the space of all functions from any Boolean algebra to
$\mathbb{R}_{\ge 0}$ together with some operations on those functions. They will
be instrumental in defining WMC as a measure in \cref{sec:wmc_as_measure} and
can be efficiently represented using ADDs. Furthermore, all of the operations
are supported by the CUDD \cite{somenzi1998cudd} package that is used by ADDMC
for ADD manipulation \cite{DBLP:conf/aaai/DudekPV20}. The definitions of
multiplication and projection are as defined by
Dudek \etal{}~\shortcite{DBLP:conf/aaai/DudekPV20}, while others are new.

@@

\section{WMC as a Measure on a Boolean Algebra} \label{sec:wmc_as_measure}

In this section, we introduce an alternative definition of WMC and demonstrate
how it relates to the standard one. Let $U$ be a set. A \emph{measure} is a
function $\mu\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ such that $\mu(\bot) = 0$,
and $\mu(a \lor b) = \mu(a) + \mu(b)$ for all $a, b \in 2^{2^U}$ whenever $a
\land b = \bot$ \cite{gaifman1964concerning,DBLP:books/daglib/0090259}. A
\emph{weight function} is a function $\nu\colon 2^U \to \mathbb{R}_{\ge 0}$. A
weight function is \emph{factored} if $\nu = \prod_{x \in U} \nu_x$ for some
functions $\nu_x\colon 2^{\{x\}} \to \mathbb{R}_{\ge 0}$, $x \in U$. We say that
a weight function $\nu\colon 2^U \to \mathbb{R}_{\ge 0}$ \emph{induces} a
measure $\mu_\nu\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ if $\mu_\nu(x) =
\sum_{\{u\} \le x} \nu(u)$.

@@

\begin{example}
  Let $U = \{a, b\}$ be a set of atoms and $\mu\colon 2^{2^U} \to
  \mathbb{R}_{\ge 0}$ a measure defined as $\mu(a \land b) = 0.72$, $\mu(a \land
  \neg b) = 0.18$, $\mu(\neg a \land b) = 0.07$, $\mu(\neg a \land \neg b) =
  0.03$.\footnote{The value of $\mu$ on any other element of $2^{2^U}$ can be
    deduced from the definition of a measure.} If $\mu$ could be represented
  using literal-weight (factored) WMC, we would have to find two weight
  functions $\nu_a\colon 2^{\{a\}} \to \mathbb{R}_{\ge 0}$ and $\nu_b\colon
  2^{\{b\}} \to \mathbb{R}_{\ge 0}$ such that $\nu = \nu_a \cdot \nu_b$ induces
  $\mu$, i.e., $\nu_a$ and $\nu_b$ would have to satisfy this system of
  equations:
  \begin{align*}
    \nu_a(\{a\}) \cdot \nu_b(\{b\}) &= 0.72 \\
    \nu_a(\{a\}) \cdot \nu_b(\emptyset) &= 0.18 \\
    \nu_a(\emptyset) \cdot \nu_b(\{b\}) &= 0.07 \\
    \nu_a(\emptyset) \cdot \nu_b(\emptyset) &= 0.03,
  \end{align*}
  which has no solutions.

@@

\section{Encoding Bayesian Networks Using Conditional Weights} \label{sec:bns}

@@

A Bayesian network is a directed acyclic graph with random variables as vertices
that defines a probability distribution over them. Let $\mathcal{V}$ denote this
set of random variables. For any random variable $X \in \mathcal{V}$, let $\im
X$ denote its set of values and $\mathrm{pa}(X)$ its set of parents. The
full probability distribution is then equal to $\prod_{X \in \mathcal{V}} \Pr(X
\mid \mathrm{pa}(X))$ \cite{DBLP:books/daglib/0023820}. For discrete Bayesian
networks (and we only consider discrete networks in this paper), each factor of
this product can be represented by a CPT. See \cref{fig:example_bn} for an
example Bayesian network that we will refer to throughout this section. For
this network, $\mathcal{V} = \{ W, F, T \}$, $\mathrm{pa}(W) = \emptyset$,
$\mathrm{pa}(F) = \mathrm{pa}(T) = \{ W \}$, $\im W = \im F = \{0, 1 \}$, and
$\im T = \{ l, m, h \}$.

@@

\subsection{Changes to ADDMC}

@@

First, ADDMC constructs the \emph{Gaifman graph} \cite{gaifman1982local} of the
input CNF formula as an aid for the algorithm's heuristics. This graph has as
vertices the variables of the formula, and there is an edge between two
variables $u$ and $v$ if there is a clause in the formula that contains both $u$
and $v$. We extend this definition to functions on Boolean algebras, i.e., the
factors of $\phi$. For any pair of distinct variables $u, v \in U$, we draw an
edge between them in the Gaifman graph if there is a function $\alpha\colon 2^X
\to \mathbb{R}_{\ge 0}$ that is a factor of $\phi$ such that $u \in X$ and $v
\in X$. For instance, a factor such as $\mathrm{CPT}_X$ will enable edges
between all distinct pairs of variables in $\mathcal{E}^*(X)$.

@@

\section{Experimental Comparison} \label{sec:experiments}

@@

\begin{figure} % TODO: update
  \centering
  \input{../paper1/cumulative.tex}%
  \caption{Cumulative number of instances solved by ADDMC over time using each
    encoding}
  \label{fig:cumulative}
\end{figure}

@@

\begin{figure*} % TODO: update
  \centering
  \input{../paper1/scatter.tex}%
  \caption{ADDMC inference time using \texttt{cw} compared to \texttt{d02}
    (left) and \texttt{sbk05} (right) for each instance in all data sets.}
  \label{fig:scatter}
\end{figure*}

According to the cumulative plot in \cref{fig:cumulative}, \texttt{cw} is
consistently able to solve more instances with any given time limit than any
other encoding, while the order of the remaining encodings from best to worst is
exactly the opposite of that in previous literature (cf. \cref{sec:related}).
Specifically, \texttt{cw} solves the same number of instances in
\SI{8.144}{\second} as \texttt{d02} does in \SI{1000}{\second}, which is the
second-best encoding in our experiments. Furthermore, \cref{tbl:tallies} shows
that \texttt{cw} was the fastest in 935 out of the 980 (i.e.,
\SI{95.4}{\percent}) instances that it managed to solve. Finally, according to
the scatter plots in \cref{fig:scatter}, \texttt{cw} is particularly promising
on Grid networks while losing to \texttt{d02} on some instances of the
Mastermind data set. We conjecture that this is so either because the particular
structure of the problem confuses the heuristics used by ADDMC or because these
instances allow for significant simplifications that are exploited by
\texttt{d02} but not \texttt{cw}.

@@

While we chose to use ADDMC \cite{DBLP:conf/aaai/DudekPV20} as the WMC algorithm
and Bayesian networks as a canonical example of a probabilistic inference task,
these are only examples meant to illustrate the broader idea that choosing a
more expressive representation of weights can outperform increasing the size of
the problem to keep the weights simple. Indeed, in this work, we have provided a
new theoretical perspective on the expressive power of WMC and illustrated the
empirical benefits of that perspective. The same idea can be adapted to other
inference problem domains such as probabilistic programs
\cite{DBLP:journals/tplp/FierensBRSGTJR15,DBLP:journals/corr/abs-2005-09089} as
well as to search-based solvers such as Cachet \cite{DBLP:conf/sat/SangBBKP04},
although the extension may be less straightforward for other WMC techniques.

%An additional benefit of conditional weights is that a more expressive format
%makes it easier to invent new WMC encodings.
%Filtering out ADDs that have nothing to do with the answer helps
%  tremendously, but I'm purposefully not doing that.

>>>>

\section{Introduction}
% TODO: mention more recent papers
% TODO: emphasise the contributions (even more)
% TODO: write ADDMC, Ace, query-dnnf, Cachet, bn2cnf, c2d in a tt font
% TODO: double-check the capitalisation of, e.g., c2d
% TODO: relate non-factorable measures to VCSPs, provenance/probabilistic
% databases
% TODO: simplify the introductory explanations (take inspiration from the
% comments of reviewer #5) (this is more about the next two sections)

% Over the last fifteen years, WMC has been extended and generalised in many
% ways, e.g., to handle continuous probability distributions
% \cite{DBLP:conf/ijcai/BellePB15}, first-order probabilistic theories
% \cite{DBLP:conf/ijcai/BroeckTMDR11,DBLP:journals/cacm/GogateD16}, and infinite
% domains \cite{DBLP:conf/aaai/Belle17}.

% , whereas approximate solvers work by sampling
% \cite{DBLP:conf/aaai/ChakrabortyFMSV14} and performing local search
% \cite{DBLP:conf/sat/WeiS05}.

% TODO: rephrase
Weighted model counting (WMC), i.e., an extension of model counting ($\#\SAT{}$)
that assigns a weight to every model \cite{DBLP:conf/aaai/SangBK05}, has emerged
as one of the most dominant and competitive approaches for handling inference
tasks in a wide range of formalisms including Bayesian networks
\cite{DBLP:conf/aaai/SangBK05} and probabilistic programs
\cite{DBLP:journals/tplp/FierensBRSGTJR15}. Furthermore, by generalising the
notion of weights to an arbitrary semiring, a range of other problems are also
captured \cite{DBLP:journals/japll/KimmigBR17}. Exact WMC solvers typically rely
on either knowledge compilation \cite{DBLP:conf/ijcai/LagniezM17} or exhaustive
DPLL search \cite{DBLP:conf/aaai/SangBK05}.

@@

% TODO: maybe summarise this in one sentence
% \paragraph{ADDs and their use in inference.} ADDs provide an efficient way to
% manipulate functions from a Boolean algebra to any algebraic structure (most
% commonly real numbers) \cite{DBLP:journals/fmsd/BaharFGHMPS97}. They have been
% used to represent value functions in Markov decision processes (MDPs)
% \cite{DBLP:conf/uai/HoeySHB99} and extended to handle first-order MDPs
% \cite{DBLP:journals/ai/SannerB09} and discrete and continuous state MDPs
% \cite{DBLP:conf/uai/SannerDB11}. The latter extension has also been used for
% weighted model integration---an extension of WMC that can handle continuous
% variables \cite{DBLP:conf/ijcai/KolbMSBK18}. For Bayesian networks, ADDS have
% been used to represent each conditional probability table (CPT) as an ADD
% \cite{DBLP:conf/icml/ZhaoMP15} and as a combination of ADDs and arithmetic
% circuits \cite{DBLP:conf/ijcai/ChaviraD07}. ADDs are particularly advantageous
% in this situation because of their ability to fully exploit context-specific
% independence, i.e., the observable structure within a CPT that is not inherited
% from the structure of the Bayesian network \cite{DBLP:conf/uai/BoutilierFGK96}.

@@

% TODO: add a link to CUDD
We also consider the space of all functions from any Boolean algebra to
$\mathbb{R}_{\ge 0}$ together with some operations on those functions. They will
be instrumental in defining WMC as a measure in \cref{sec:wmc_as_measure} and
can be efficiently represented using ADDs. Furthermore, all of the operations
are supported by CUDD package that is used by ADDMC for ADD manipulation
\cite{DBLP:conf/aaai/DudekPV20}. The definitions of multiplication and
projection are as defined by Dudek \etal{}~\shortcite{DBLP:conf/aaai/DudekPV20},
while others are new.

@@

\section{WMC as a Measure on a Boolean Algebra} \label{sec:wmc_as_measure}
% TODO beginning of section 4: a weight function is *any* function; end the
% paragraph with 'for all x \in U'.

% TODO: mention that putting a measure on a BA is a well-known idea
In this section, we introduce an alternative definition of WMC and demonstrate
how it relates to the standard one. Let $U$ be a set. A \emph{measure} is a
function $\mu\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ such that $\mu(\bot) = 0$,
and $\mu(a \lor b) = \mu(a) + \mu(b)$ for all $a, b \in 2^{2^U}$ whenever $a
\land b = \bot$. A \emph{weight function} is a function $\nu\colon 2^U \to
\mathbb{R}_{\ge 0}$. A weight function is \emph{factored} if $\nu = \prod_{x \in
  U} \nu_x$ for some functions $\nu_x\colon 2^{\{x\}} \to \mathbb{R}_{\ge 0}$,
$x \in U$. We say that a weight function $\nu\colon 2^U \to \mathbb{R}_{\ge 0}$
\emph{induces} a measure $\mu_\nu\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ if
$\mu_\nu(x) = \sum_{\{u\} \le x} \nu(u)$.

@@

\begin{example}
  Let $U = \{a, b\}$ be a set of atoms and $\mu\colon 2^{2^U} \to
  \mathbb{R}_{\ge 0}$ a measure defined as $\mu(a \land b) = 0.72$, $\mu(a \land
  \neg b) = 0.18$, $\mu(\neg a \land b) = 0.07$, $\mu(\neg a \land \neg b) =
  0.03$.\footnote{The value of $\mu$ on any other element of $2^{2^U}$ can be
    deduced from the definition of a measure.} If $\mu$ could be represented
  using literal-weight (factored) WMC, we would have to find two weight
  functions $\nu_a\colon 2^{\{a\}} \to \mathbb{R}_{\ge 0}$ and $\nu_b\colon
  2^{\{b\}} \to \mathbb{R}_{\ge 0}$ such that $\nu = \nu_a \cdot \nu_b$ induces
  $\mu$, i.e., $\nu_a$ and $\nu_b$ would have to satisfy this system of
  equations:
  \begin{align*}
    \nu_a(\{a\}) \cdot \nu_b(\{b\}) &= 0.72 \\
    \nu_a(\{a\}) \cdot \nu_b(\emptyset) &= 0.18 \\
    \nu_a(\emptyset) \cdot \nu_b(\{b\}) &= 0.07 \\
    \nu_a(\emptyset) \cdot \nu_b(\emptyset) &= 0.03,
  \end{align*}
  which has no solution.

@@

\section{Encoding Bayesian Networks Using Conditional Weights} \label{sec:bns}
% TODO: make it clear which parts of the encoding are not new
% TODO: rewrite the encoding without using an algorithm (look into the feedback
% of reviewer #5 and someone before him)
% TODO: explain/prove why it's impossible for the all-zero encoding to represent
% the last value (a.k.a. the -implicit flag)

@@

A Bayesian network is a directed acyclic graph with random variables as vertices
that defines a probability distribution over them. Let $\mathcal{V}$ denote this
set of random variables. For any random variable $X \in \mathcal{V}$, let $\im
X$ denote its set of values and $\mathrm{pa}(X)$ its set of parents. The
full probability distribution is then equal to $\prod_{X \in \mathcal{V}} \Pr(X
\mid \mathrm{pa}(X))$. For discrete Bayesian networks (and we only consider
discrete networks in this paper), each factor of this product can be represented
by a CPT. See \cref{fig:example_bn} for an example Bayesian network that we will
refer to throughout this section. For this network, $\mathcal{V} = \{ W, F, T
\}$, $\mathrm{pa}(W) = \emptyset$, $\mathrm{pa}(F) = \mathrm{pa}(T) = \{ W \}$,
$\im W = \im F = \{0, 1 \}$, and $\im T = \{ l, m, h \}$.

@@

\subsection{Changes to ADDMC}
% TODO: clarify that both variable orders are handled by the ADDMC heuristics

@@

% TODO: Gaifman graph is also known as the primal graph
First, ADDMC constructs the \emph{Gaifman graph} of the input CNF formula as an
aid for the algorithm's heuristics. This graph has as vertices the variables of
the formula, and there is an edge between two variables $u$ and $v$ if there is
a clause in the formula that contains both $u$ and $v$. We extend this
definition to functions on Boolean algebras, i.e., the factors of $\phi$. For
any pair of distinct variables $u, v \in U$, we draw an edge between them in the
Gaifman graph if there is a function $\alpha\colon 2^X \to \mathbb{R}_{\ge 0}$
that is a factor of $\phi$ such that $u \in X$ and $v \in X$. For instance, a
factor such as $\mathrm{CPT}_X$ will enable edges between all distinct pairs of
variables in $\mathcal{E}^*(X)$.

@@

\section{Experimental Comparison} \label{sec:experiments}
% TODO: stacked bar charts: 1) mention that the error bars denote 0.1 * standard
% deviation
% 2) and that only successfully run instances are included
% 3) and that the top error bar relates only to inference (and not the sum of
% the two)
% TODO: note: I'm unable to replicate the experiments from bklm. This could be
% due to two reasons:
% 1) I'm measuring running time rather than compiled structure size <- more
% likely
% 2) I'm not using the -implicit flag.
% TODO: We use Ace to compile the BN together with its evidence (as opposed to
% compiling just the BN and using evidence during inference) as this more
% closely resembles the behaviour of ADDMC
% TODO: we run cd05 only with dtBnMinfill, as that's the more commonly used
% setting (although another one was used as well)
% TODO: The -implicit flag of bklm16 cannot be used because even a small
% Bayesian network with seven independent binary variables with 0.1/0.9
% probabilities is already big enough to drive the scaling factor down to zero.
% The last step of this encoding is a multiplication of a huge number (on the
% order of 10^30 for some of the smaller DQMR instances) by a number
% equivalently close to zero (so that their product lands in the interval [0,
% 1]). Thus, we disable this flag (thus making the algorithm slower but
% correct).
% TODO: the legacy mode encoding of bklm16 also runs c2d (as packaged with Ace)
% TODO: query-dnnf: http://www.cril.univ-artois.fr/kc/d-DNNF-reasoner.html
% TODO: bn2cnf http://www.cril.univ-artois.fr/KC/bn2cnf.html
% TODO: Note that our encoder was implemented in Python with no concern for
% efficiency, so encoding could easily be improved by one or two orders of
% magnitude.
% TODO: Python, Java, gcc versions (maybe also all the software that I'm using
% (e.g., bn2cnf, Cachet))
% TODO: 85% is as recommended in the README of Ace (might have to go lower than that)
% TODO: for each encoding, mention:
% 1) software
% 2) encoding options
% 3) additional data processing
% 4) inference options

@@

\begin{figure}
  \centering
  \input{../paper1/cumulative.tex}%
  \caption{Cumulative number of instances solved by ADDMC over time using each
    encoding}
  \label{fig:cumulative}
\end{figure}

@@

\begin{figure*}
  \centering
  \input{../paper1/scatter.tex}%
  \caption{ADDMC inference time using \texttt{cw} compared to \texttt{d02}
    (left) and \texttt{sbk05} (right) for each instance in all data sets.}
  \label{fig:scatter}
\end{figure*}

% TODO: Either remove the 'conjecture' (preferable), or explain why those
% instances perform worse
According to the cumulative plot in \cref{fig:cumulative}, \texttt{cw} is
consistently able to solve more instances with any given time limit than any
other encoding, while the order of the remaining encodings from best to worst is
exactly the opposite of that in previous literature (cf. \cref{sec:related}).
Specifically, \texttt{cw} solves the same number of instances in
\SI{8.144}{\second} as \texttt{d02} does in \SI{1000}{\second}, which is the
second-best encoding in our experiments. Furthermore, \cref{tbl:tallies} shows
that \texttt{cw} was the fastest in 935 out of the 980 (i.e.,
\SI{95.4}{\percent}) instances that it managed to solve. Finally, according to
the scatter plots in \cref{fig:scatter}, \texttt{cw} is particularly promising
on Grid networks while losing to \texttt{d02} on some instances of the
Mastermind data set. We conjecture that this is so either because the particular
structure of the problem confuses the heuristics used by ADDMC or because these
instances allow for significant simplifications that are exploited by
\texttt{d02} but not \texttt{cw}.

@@

% TODO: clarify what 'more expressive' means
While we chose to use ADDMC \cite{DBLP:conf/aaai/DudekPV20} as the WMC algorithm
and Bayesian networks as a canonical example of a probabilistic inference task,
these are only examples meant to illustrate the broader idea that choosing a
more expressive representation of weights can outperform increasing the size of
the problem to keep the weights simple. Indeed, in this work, we have provided a
new theoretical perspective on the expressive power of WMC and illustrated the
empirical benefits of that perspective. The same idea can be adapted to other
inference problem domains such as probabilistic programs
\cite{DBLP:journals/tplp/FierensBRSGTJR15} as
well as to search-based solvers such as Cachet \cite{DBLP:conf/sat/SangBBKP04},
although the extension may be less straightforward for other WMC techniques.

% TODO: emphasise that the simplicity of the BN encoding implies the strength of
% the general principle.
% An additional benefit of conditional weights is that a more expressive format
% makes it easier to invent new WMC encodings.
% Filtering out ADDs that have nothing to do with the answer helps
%  tremendously, but I'm purposefully not doing that.
